---
title: "Length based indices via mar"
---

```{r include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE,
                      warning = FALSE)
```

**NOTE**: After the lecture today, adjusted code to take into account stations with zero catch. The code only works for one species at the time.

**TODO**: There is some bug related to the 2009 data calculation that needs fixing.

# Preamble
___

This case example demonstrates how to calculate standardized survey indices from SMB using the tidyverse approach.

In principle the method is described in [Fjölrit 131](https://www.hafogvatn.is/static/research/files/fjolrit-131pdf). The pseudo-code is something like:
```
1.  get survey stations %>%
2.  get length data %>%
3.  get count data %>% 
4.  scale by counted %>% 
5.  trim towlength %>%                  # narrow extremes
6.  standardize by towlength %>%        # on the number only
7.  calculate_biomass %>%               # up to now one recort per length class
8.  summarise by station %>%            # one record per station
9.  filter stations                     # fixed, e.g. tognumer %in% 1:39
10. summarise by strata %>%             # one record per strata
11. raise to strata area %>%            # 
12. summarise by year                   # one record per year
```

# Setup
___

```{r}
library(tidyverse)
library(mar)
con <- connect_mar()
```


## Additonal tables in mar

In order to calculate standardized survey indices within Oracle we need in addition to the tables in mar Oracle tables on:

* Length-weight coefficients
* What strata each tow belongs to
* The area of the strata

For this case example these tables have been made available to all:
```{r}
glimpse(tbl_mar(con, "ops$einarhj.lwcoeff"))
glimpse(tbl_mar(con, "ops$einarhj.smbstationsstrata"))
glimpse(tbl_mar(con, "ops$einarhj.oldstrataarea"))
```

The smbstationsstrata just matches synis_id (tow) with a particular stratanumber. The oldstrataarea contains the area (in km2) for each strata.

If one is not interested in running the code within Oracle one would need to supply the above tables as an r-dataframe. The side effect, one would need to change the code below such that anytime an Oracle-table is addressed one would need to add the `collect`-function.

## Select a species

Lets select a species to work with and the length above which we want to obtain
the standardized survey indices:
```{r}
Species <- 23                     # Select a species
Length.min <- 5                   # Minimum length for indices calculation
Length.max <- 500                 # Maximum length for indices calculation
```

## Some constants

Some constants used in the standardization process (you can deviate from them if you want):
```{r}
std.cv        <- 1             # The cv if only one station in a strata
std.towlength <- 4             # Standard tow length is 4 nautical miles
std.width     <- 17            # Standard sweep width in meters
# Standard area swept, units of square nautical miles
#   (1852 meters / nm)
std.area <- std.towlength * std.width / 1852

min.towlength <- 2             # Minimum "acceptable" towlength
max.towlength <- 8             # Maximum "acceptable" towlength
```

A little helper function:
```{r}
#' Calculate overall cv from stratfied summary statistics
#' 
#' @param m Mean value within strata
#' @param s Standard deviation within strata
#' @param area The area (e.g. survey strata area)
#' @param n number of samples within strata
#'
#' @export
#'
calc_cv <- function(m, s, area, n) {
  
  Mean = sum(m * area) / sum(area)
  Sum = sum(m * area)
  tmpsum = sum(m[!is.na(s)] * area[!is.na(s)])
  Calc.sdev = sqrt(sum(s[!is.na(s)]^2 * area[!is.na(s)]^2/  n[!is.na(s)])   / sum(area[!is.na(s)])^2)
  Sdev = Calc.sdev * Sum/tmpsum
  cv = Sdev/Mean
  
  return(cv)
}
```

# The code
___

```{r}
d <-  
  # 1. get survey stations -----------------------------------------------------
  lesa_stodvar(con) %>% 
  filter(synaflokkur == 30, veidarfaeri == 73) %>% 
  select(synis_id, ar, toglengd, reitur, tognumer) %>% 
  
  # 2. get length data ---------------------------------------------------------
  left_join(lesa_lengdir(con) %>% 
              filter(tegund %in% Species,
                     lengd >= Length.min,
                     lengd < Length.max) %>% 
              group_by(synis_id, tegund, lengd) %>% 
              summarise(fjoldi = sum(fjoldi, na.rm = TRUE))) %>% 
              ungroup() %>% 
  
  # 3. get count data ----------------------------------------------------------
  left_join(lesa_numer(con) %>% 
              mutate(r =  ifelse(fj_maelt != 0, 1 + fj_talid/fj_maelt, 1)) %>% 
              select(synis_id, tegund, r)) %>% 
  
  # 0. A temporary fix, for zero stations --------------------------------------
  #     TODO: Find a more permanent solution so scripts works for more than
  #           one species (via group_by( ..., tegund))
  mutate(tegund = if_else(is.na(tegund), Species, tegund),
         lengd  = if_else(is.na(tegund), Length.min, lengd),
         fjoldi = if_else(is.na(tegund), 0, fjoldi),
         r      = if_else(is.na(tegund), 1, r)) %>% 
  
  # 4. scale by counted --------------------------------------------------------
  mutate(N = r * fjoldi / 1e3) %>%   # units of thousand
  
  # 5. trim towlength ----------------------------------------------------------
  mutate(toglengd = if_else(toglengd > max.towlength, max.towlength, toglengd),
         toglengd = if_else(toglengd < min.towlength, min.towlength, toglengd)) %>% 
  
  # 6. standardize by towlength ------------------------------------------------
  mutate(N = N * std.towlength/toglengd)  %>% # standardize to per 4 miles
  
  # 7. calculate_biomass from numbers, length and a and b ----------------------
  # 7.a get the length weight coefficients
  left_join(tbl_mar(con, "ops$einarhj.lwcoeff")) %>% 
  # 7.b use Newton's law if lwcoefficient for species not specified
  mutate(a = ifelse(is.na(a), 0.01, a),
         b = ifelse(is.na(b), 3.00, b),
         B  = ifelse(is.na(N), 0, N) * a * lengd^b/1e3) %>% 
  
  # 8. summarise by station ----------------------------------------------------
  # NOTE: here is the first step where statistics by length is dropped
  #       some (minor) recoding above would be needed if one were to take things
  #       forward by each length class
  group_by(synis_id, tognumer, ar) %>% 
  summarise(N = sum(N, na.rm = TRUE),
            B = sum(B, na.rm = TRUE)) %>% 
  # Zero stations
  mutate(N = ifelse(is.na(N), 0, N),
         B = ifelse(is.na(B), 0, B)) %>% 
  
  # 9. filter stations ---------------------------------------------------------  
  filter(tognumer %in% 1:39 | is.na(tognumer)) %>%
  
  # 10. summarise by strata ----------------------------------------------------
  # 10.a  Get the strata for each station
  left_join(tbl_mar(con, "ops$einarhj.smbstationsstrata") %>% 
              select(synis_id, strata = oldstrata)) %>% 
  # 10.b group by year and strata and calculate number of stations, mean and sd
  group_by(ar, strata) %>% 
  summarise(sN  = n(),   # number of stations within strata
            n_m  = mean(N, na.rm = TRUE),
            n_d  = ifelse(n() == 1, mean(N, na.rm = TRUE) * std.cv, sd(N)),
            b_m  = mean(B, na.rm = TRUE),
            b_d  = ifelse(n() == 1, mean(B, na.rm = TRUE) * std.cv, sd(B))) %>% 
  
  # 11. raise to strata area ---------------------------------------------------
  # 11.a get area of the strata
  left_join(tbl_mar(con, "ops$einarhj.oldstrataarea") %>% 
              select(strata = oldstrata, area = rall.area)) %>% 
  # 11.b do the strata raising
  #  area is above is in km2, here convert nm2
  mutate(area  = area/1.852^2 / std.area) %>% 
  mutate(n     = n_m  * area,
         b     = b_m  * area) %>% 
  
  # ----------------------------------------------------------------------------
  # Up to now we only have been operating within Oracle. I.e. sql-scripts via R.
  # Have to collect here because of calc_cv function in the year aggregate step
  # TODO: Fix that, do internally in Oracle
  collect(n = Inf) %>% 
  
  # 11. summarise by year ------------------------------------------------------
  group_by(ar) %>%
  summarise(n = sum(n, na.rm = TRUE),
            # A la Höski
            n.cv = calc_cv(n_m, n_d, area, sN),
            b = sum(b, na.rm = TRUE),
            # A la Höski
            b.cv = calc_cv(b_m, b_d, area, sN)) %>% 
  ungroup()
```


# The results
___

```{r}
glimpse(d)
```

So we have a dataframe where each row is year and then the abundance and biomass indices and the cv. Lets plot the biomass indices:
```{r}
d %>% 
  ggplot(aes(ar + 3/12, b)) +
  geom_point() +
  geom_linerange(aes(ymin = b * (1 - b.cv),
                      ymax = b * (1 + b.cv))) +
  expand_limits(y = 0) +
  labs(x = NULL, y = NULL,
       title = "Spring survey biomass indices")
```

NOTE: Bug in the code, no cv is returned for the year 2009.

# Meeting with the Pope

Below a comparison with results based on Pope Husky's code, just as a double check.

```{r, echo = FALSE}
attach("/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMB/Allaggroldsmbindex.rdata")
vatican <-
 Allaggroldsmbindex %>%
 filter(species == Species,
        svaedi == "Heild",
        diurnal == 0,
        fixed == 0) %>%
 select(ar, lengd, b = bio.staerri, b.cv = cv.bio.staerri) %>%
 mutate(source = "Vatican") %>%
 as_tibble()
vatican <-
  vatican %>% 
  filter(lengd == max(min(lengd), Length.min))
detach("file:/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMB/Allaggroldsmbindex.rdata")

d %>% 
  mutate(source = "tidy",
         ar = ar + 3/12) %>% 
  bind_rows(vatican) %>% 
  ggplot(aes(ar, b, colour = source)) +
  geom_point() +
  geom_pointrange(aes(ymin = b * (1 - b.cv),
                      ymax = b * (1 + b.cv))) +
  scale_colour_brewer(palette = "Set1") +
  theme(legend.position = c(0.25, 0.8)) +
  expand_limits(y = 0) +
  labs(x = NULL, y = NULL,
       title = "Biomass indices",
       subtitle = "Comparison of the orthodoxy (Vatican) and tidy")
```

Besides that tidy and the Vatican are not operating in the same time zone (one has argued that tidy operates under the Greek-orthodox time zone, but it seems inverse) the stuff on the y-axis looks reasonably similar.

If you want to take a look the Vatican code check out the following r-scripts:

```{r}
dir("/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMB", pattern = "*.[rR]$",
    full.names = TRUE)
```

The backbone script is found in:
```{r, eval = FALSE}
# Not run
parse("/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMB/Indices.r")
```

with the workhorse being the Calc.index-function.

# Homework

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

Homework **only** for those with some inside knowledge of the fall survey indices.

The fall survey stations can be filtered out of the station table by recognizing that:

* synaflokkur is 35
* gear (veidarfaeri) is 77 (shallow) and 78 (deep)
* towid (tognumer) are less than 80
* the sweep area of gear 78 is set to 1.25 times that of the shallow gear

The fall survey indices are stratified according to "newstrata", check out:
```{r, eval = FALSE}
# Not run
tbl_mar(con, "ops$einarhj.smhstationsstrata")
```

The area of the newstrata are found in:
```{r, eval = FALSE}
# Not run
tbl_mar(con, "ops$einarhj.newstrataarea")
```

Make minimum adjustment to the code to calculate the fall biomass survey indices. The resulting plot should look something like:

```{r, echo = FALSE}
# The fall survey
d <-  
  # 1. get survey stations -----------------------------------------------------
  lesa_stodvar(con) %>% 
  filter(synaflokkur == 35,
         veidarfaeri %in% c(77, 78),
         ar > 1995,
         !ar %in% 2011, 
         tognumer < 80) %>% 
  #mutate(r_sweep = if_else(veidarfaeri == 78, 1.25, 1)) %>% 
  # NOTE: in /net/hafkaldi/export/u2/reikn/R/SurveyWork/SMH/stations.r
  # Additional filters are:
  #     ar > 1995, !ar %in% 2011, tognumer < 80) %>%
  # then we have this: 
  #     mutate(mult = if_else(veidarfaeri == 78, 1.25, 1))
  #     mutate(index = reitur*100+tognumer)
  # handle that here or below below??
  # ALSO:
  #      i <- is.na(haustrall.all.st$kl.kastad)
  #      if(any(i)) haustrall.all.st$kl.kastad[i] <- 12
  #      haustrall.all.st <- inside.reg.bcbreytt(haustrall.all.st)

  select(synis_id, ar, gid = veidarfaeri, toglengd, reitur, tognumer) %>% 
  
  # 2. get length data ---------------------------------------------------------
  left_join(lesa_lengdir(con) %>% 
              filter(tegund %in% Species,
                     lengd >= Length.min,
                     lengd < Length.max) %>% 
              group_by(synis_id, tegund, lengd) %>% 
              summarise(fjoldi = sum(fjoldi, na.rm = TRUE))) %>% 
              ungroup() %>% 
  
  # 3. get count data ----------------------------------------------------------
  left_join(lesa_numer(con) %>% 
              mutate(r =  ifelse(fj_maelt != 0, 1 + fj_talid/fj_maelt, 1)) %>% 
              select(synis_id, tegund, r)) %>% 
  
  # 0. A temporary fix, for zero stations --------------------------------------
  #     TODO: Find a more permanent solution so scripts works for more than
  #           one species (via group_by( ..., tegund))
  mutate(tegund = if_else(is.na(tegund), Species, tegund),
         lengd  = if_else(is.na(tegund), Length.min, lengd),
         fjoldi = if_else(is.na(tegund), 0, fjoldi),
         r      = if_else(is.na(tegund), 1, r)) %>% 
  
  # 4. scale by counted --------------------------------------------------------
  mutate(N = r * fjoldi / 1e3) %>%  # units of thousand

  # X. scale by gear
  #  one gear has a wider sweep than the others
  mutate(N = if_else(gid == 78, 1.25, 1) * N) %>%

  
  # 5. trim towlength ----------------------------------------------------------
  mutate(toglengd = if_else(toglengd > max.towlength, max.towlength, toglengd),
         toglengd = if_else(toglengd < min.towlength, min.towlength, toglengd)) %>% 
  
  # 6. standardize by towlength ------------------------------------------------
  mutate(N = N * std.towlength/toglengd)  %>% # standardize to per 4 miles
  
  # 7. calculate_biomass from numbers, length and a and b ----------------------
  # 7.a get the length weight coefficients
  left_join(tbl_mar(con, "ops$einarhj.lwcoeff")) %>% 
  # 7.b use Newton's law if lwcoefficient for species not specified
  mutate(a = ifelse(is.na(a), 0.01, a),
         b = ifelse(is.na(b), 3.00, b),
         B  = ifelse(is.na(N), 0, N) * a * lengd^b/1e3) %>% 
  
  # 8. summarise by station ----------------------------------------------------
  # NOTE: here is the first step where statistics by length is dropped
  #       some (minor) recoding above would be needed if one were to take things
  #       forward by each length class
  group_by(synis_id, tognumer, ar) %>% 
  summarise(N = sum(N, na.rm = TRUE),
            B = sum(B, na.rm = TRUE)) %>% 
  # Zero stations
  mutate(N = ifelse(is.na(N), 0, N),
         B = ifelse(is.na(B), 0, B)) %>% 
  
  # 9. filter stations ---------------------------------------------------------  
  # filter(tognumer %in% 1:39 | is.na(tognumer)) %>%
  
  # 10. summarise by strata ----------------------------------------------------
  # 10.a  Get the strata for each station
  left_join(tbl_mar(con, "ops$einarhj.smhstationsstrata") %>% 
              select(synis_id, strata = newstrata)) %>% 
  # 10.b group by year and strata and calculate number of stations, mean and sd
  group_by(ar, strata) %>% 
  summarise(sN  = n(),   # number of stations within strata
            n_m  = mean(N, na.rm = TRUE),
            n_d  = ifelse(n() == 1, mean(N, na.rm = TRUE) * std.cv, sd(N)),
            b_m  = mean(B, na.rm = TRUE),
            b_d  = ifelse(n() == 1, mean(B, na.rm = TRUE) * std.cv, sd(B))) %>% 
  
  # 11. raise to strata area ---------------------------------------------------
  # 11.a get area of the strata
  left_join(tbl_mar(con, "ops$einarhj.newstrataarea") %>% 
              select(strata = newstrata, area = rall.area)) %>% 
  # 11.b do the strata raising
  #  area is above is in km2, here convert nm2
  mutate(area  = area/1.852^2 / std.area) %>% 
  mutate(n     = n_m  * area,
         b     = b_m  * area) %>% 
  
  # ----------------------------------------------------------------------------
  # Up to now we only have been operating within Oracle. I.e. sql-scripts via R.
  # Have to collect here because of calc_cv function in the year aggregate step
  # TODO: Fix that, do internally in Oracle
  collect(n = Inf) %>% 
  
  # 11. summarise by year ------------------------------------------------------
  group_by(ar) %>%
  summarise(n = sum(n, na.rm = TRUE),
            # A la Höski
            n.cv = calc_cv(n_m, n_d, area, sN),
            b = sum(b, na.rm = TRUE),
            # A la Höski
            b.cv = calc_cv(b_m, b_d, area, sN)) %>% 
  ungroup()

d %>% 
  ggplot(aes(ar, b)) +
  geom_point() +
  geom_pointrange(aes(ymin = b * (1 - b.cv),
                      ymax = b * (1 + b.cv))) +
  scale_colour_brewer(palette = "Set1") +
  theme(legend.position = c(0.25, 0.8)) +
  expand_limits(y = 0) +
  labs(x = NULL, y = NULL,
       title = "Biomass indices",
       subtitle = "Comparison of the orthodoxy (Vatican) and tidy")
```


```{r, echo = FALSE, eval = FALSE}
# Comparison with the Vatican
attach("/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMH/Allaggrsmhindex.rdata")
vatican <-
 Allaggrsmhindex %>%
 filter(species == Species,
        svaedi == "Heild",
        diurnal == 0,
        fixed == 0,
        skipShallow == 0) %>%
 select(ar, lengd, b = bio.staerri, b.cv = cv.bio.staerri) %>%
 mutate(source = "Vatican") %>%
 as_tibble()
vatican <-
  vatican %>% 
  filter(lengd == max(min(lengd), Length.min))
detach("file:/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMH/Allaggrsmhindex.rdata")

d %>% 
  mutate(source = "tidy",
         ar = ar + 3/12) %>% 
  bind_rows(vatican) %>% 
  ggplot(aes(ar, b, colour = source)) +
  geom_point() +
  geom_pointrange(aes(ymin = b * (1 - b.cv),
                      ymax = b * (1 + b.cv))) +
  scale_colour_brewer(palette = "Set1") +
  theme(legend.position = c(0.25, 0.8)) +
  expand_limits(y = 0) +
  labs(x = NULL, y = NULL,
       title = "Biomass indices",
       subtitle = "Comparison of the orthodoxy (Vatican) and tidy")
```

NOTE: Bug in the code, no cv returned for year 2017

</div>
</div>

# Appendix

Documentation of the code used to generate the additional tables in mar.

## Old strata

```{r, eval = FALSE}
attach("/net/hafkaldi/export/u2/reikn/R/SurveyWork/OldStratas/.RData")
d <- 
  data_frame(oldstrata = attributes(STRATAS)$names %>% as.integer(),
             area = attributes(STRATAS)$area,
             rall.area = attributes(STRATAS)$rall.area)
dbWriteTable(con, name = "OLDSTRATAAREA", value = d, overwrite = TRUE)
detach("file:/net/hafkaldi/export/u2/reikn/R/SurveyWork/OldStratas/.RData")
# sql: grant select on oldstrataarea to h_fiskar_skoda
```

## New strata

```{r, eval = FALSE}
attach("/net/hafkaldi/export/u2/reikn/R/SurveyWork/NewStratas/.RData")
d <- 
  data_frame(newstrata = 1:length(attributes(STRATAS)$names),
             area = attributes(STRATAS)$area,
             rall.area = attributes(STRATAS)$rall.area)
dbWriteTable(con, name = "NEWSTRATAAREA", value = d, overwrite = FALSE)
detach("file:/net/hafkaldi/export/u2/reikn/R/SurveyWork/NewStratas/.RData")
# sql: grant select on newstrataarea to h_fiskar_skoda
```

## SMB

```{r, eval = FALSE}
attach("/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMB/Stations.rdata")
d <- 
  STODVAR.all %>%
  select(synis_id = synis.id, oldstrata, newstrata)
dbWriteTable(con, name = "SMBSTATIONSSTRATA", value = d, overwrite = TRUE)
detach("file:/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMB/Stations.rdata")
# sql: grant select on smbstationsstrata to h_fiskar_skoda
```

## SMH

```{r, eval = FALSE}
attach("/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMH/stations.rdata")
d <- 
  haustrall.all.st %>%
  select(synis_id = synis.id, newstrata) %>% 
  mutate(newstrata = as.integer(newstrata))
dbWriteTable(con, name = "SMHSTATIONSSTRATA", value = d, overwrite = FALSE)
detach("file:/net/hafkaldi/export/u2/reikn/R/SurveyWork/SMH/stations.rdata")
# sql: grant select on smhstationsstrata to h_fiskar_skoda
```



