---
title: "Spatial data and ggplot"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
```

# Preamble

```{r, message = FALSE, warning = FALSE, results='hide'}
# devtools::install_github("tidyverse/ggplot2")  # has geom_sf
library(tidyverse)
library(lubridate)
library(mar)

# use install.packages("packagename") if packages below are not
#   already on your computer
library(geo)               # the old faithful
library(maps)              # basic low resolution maps
library(mapdata)           # some high resolution maps
library(marmap)            # to get bathymetric data from NOAA or GEBCO
library(ggmap)             # tile backgrounds
library(sp)                # spatial stuff
library(rgdal)             # to read in ESRI files
```

Lets continue using the minke data, because it has two columns that refer to geographical coordinates. Lets plot the "kill location" of the minkes: 
```{r}
minke <- read_csv("ftp://ftp.hafro.is/pub/tcrenv/minke.csv")
minke %>% 
  ggplot(aes(lon, lat)) +
  geom_point(colour = "red")
```

In the above plot, we basically have mapped the longitude on the x-axis and the latitude on the y-axis. There are two things missing:

* A background, giving the reader a better indication of the geographical region of the sample location.
* The projection (aspect ratio between the x- and the y-axis) looks wrong

Lets add some random island as a background and use the "Mercator" projection (the default):

```{r}
minke %>% 
  ggplot(aes(lon, lat)) +
  geom_path(data = geo::island) +   # so, this object is in the geo-package
  geom_point(colour = "red") +
  coord_map(projection = "mercator")
```

Here "the background map" is just a dataframe with a set of longitudinal and latitudinal coordinates (check out 'glimpse(island)') that are **arrange**-ed in a specific way. And the function `geom_path` respects that order (try to use `geom_line` instead in the above code). The call to `coord_map` in its most simple of terms sets the proper aspect ratio between the x- and the y-axis.

# On backgrounds

## Iceland

A more detail map of shorelines than provided in the geo-package (besides the bisland-object) can be obtained from [LMI](http://www.lmi.is). A thinned version has been conveniently placed at ftp://ftp.hafro.is/pub/reiknid/einar/shapes in the form of a zipped ESRI shapefiles (iceland.zip). To read it into R one would need to download the bundle first:

```{r, eval = FALSE}
download.file("ftp://ftp.hafro.is/pub/reiknid/einar/shapes/iceland.zip",
              destfile = "data/iceland.zip")
unzip("data/iceland.zip", exdir = "data")
```

Lets see what we have got:
```{r}
dir("data", pattern = "iceland")
```

So we have, besides the .zip file four files ending with .dbf, .prj, .shp and .shx. These are standard extensions of [ESRI shapefiles](https://en.wikipedia.org/wiki/Shapefile)

And then read it into R with the `readOGR` function that resides in the rgdal-package:
```{r}
iceland <- 
  rgdal::readOGR("data/iceland.shp", verbose = FALSE)
class(iceland)
# str(iceland)
```

The class of the object iceland is "SpatialPolygonsDataFrame" and the structure is kind of obtuse. Let's not worry too much about that now because `ggplot` can operate on these classes directly (example not shown here). What really happens behind the scene though is that the object is turned into a dataframe via the `fortify`-function. Let's look at what that returns:

```{r}
iceland <- iceland %>% fortify()
glimpse(iceland)
```

Here the coordinates are stored in columns "long" and "lat" with each "island" separated by values in the variable "group". Hence to plot it we use the 'group' argument in the `aes`-function:

```{r}
ggplot() +
  geom_polygon(data = iceland, aes(long, lat, group = group)) +
  geom_point(data = minke, aes(lon, lat), col = "red") +
  coord_map(xlim = c(-22.8, -21.7), ylim = c(63.95, 64.4))
```

Here there are two additional element introduced:

* The **variable** group in the iceland-dataframe is a unique identifier of each separate shape (islands, scerries, etc.). By specifying the **argument** group in the `aes`-function one prevents that polygons are drawn across separate elements.
* The limits are specified inside the call in function `coord_quickmap`. This allows us to zoom into a specific region (try `coord_map` without any argument).

If we had not constructed the code as above we would have gotten a bit of a mess. Try:
```{r eval = FALSE}
ggplot() +
  geom_polygon(data = iceland, aes(long, lat)) +
  coord_map(xlim = c(-22.8, -21.7), ylim = c(63.95, 64.4))
ggplot() +
  geom_polygon(data = iceland, aes(long, lat, group = group)) +
  coord_map() +
  xlim(c(-22.8, -21.7)) + ylim(c(63.95, 64.4))
```

The original shapefile from LMI, which has a considerable higher resolution is available (at least temporarily) in the "biceland.zip" bundle in the [ftp-directory](ftp://ftp.hafro.is/pub/reiknid/einar/shapes)

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

As an exercise in downloading, extracting, importing and plotting a shapefile try to plot the minke data over the bormicon-area shape file (although not really a shoreline), the location of the zipped bundle being ftp://ftp.hafro.is/pub/reiknid/einar/shapes/bormicon.zip

```{r, echo = FALSE, eval = FALSE}
download.file("ftp://ftp.hafro.is/pub/reiknid/einar/shapes/bormicon.zip",
              destfile = "data/bormicon.zip")
unzip("data/bormicon.zip", exdir = "data")
rgdal::readOGR("data/bormicon.shp", verbose = FALSE) %>% 
  ggplot() +
  geom_path(aes(long, lat, group = group)) +
  geom_point(data = minke, aes(lon, lat), col = "red") +
  coord_quickmap()
```

</div>
</div>

## Global shorelines

Some few of you may be working on a larger scale than the some random islands. The country/shorelines accessible via function in the map-package may be enough to serve your needs.

```{r}
m <- map_data("world")
ggplot(m, aes(long, lat, group = group)) +
  geom_polygon() +
  coord_quickmap()
```

A little higher resolution of the global country profiles is available in the mapdata-package. E.g. to get a map of the North Sea one could do:

```{r}
xlim <- c(-5, 12)
ylim <- c(50, 60)
m <- map_data("worldHires", xlim = xlim, ylim = ylim)
ggplot(m, aes(long, lat, group = group)) +
  geom_polygon() +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE)
```

#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Homework</div>
<div class="panel-body">

1. Play around by selecting and plotting different regions or areas
2. Read the `coord_map` help file for projections other than the default "Mercator". Try to create a map with different projections, e.g. one that looks something along (hint: ortho):

```{r, echo = FALSE}
map_data("world") %>% 
  ggplot(aes(x = long, y = lat, group = group)) +
  theme_bw() +
  geom_polygon(fill = "grey") +
  scale_y_continuous(NULL) +
  scale_x_continuous(NULL) +
  coord_map("ortho", xlim = c(-45,30), ylim = c(35,90))
```

</div>
</div>

## Background images

The ggmap-packages allows one to obtain background images from the web via the `get_map`-function:

```{r}
m2 <- get_map(location = c(-19,65), zoom= 6)
m2 <- ggmap(m2) +
  labs(x = NULL, y = NULL)
m2
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Read the help file for `get_map` and get at minimum a satellite maptype.
2. Overlay the minke data
3. Try to limit the plot range to the range of the minke data


```{r, echo = FALSE, eval = FALSE}
m2 <- get_map(location = c(-19,65), zoom= 6, maptype = "satellite")
m2 %>% 
  ggmap() +
  geom_point(data = minke, aes(lon, lat), colour = "red") +
  labs(x = NULL, y = NULL) +
  xlim(range(minke$lon)) +    # limit the map range  
  ylim(range(minke$lat))      # to the data range
```

</div>
</div>

Although using background "images" are impressive, the novelty wears off quickly. Think carefully about not cluttering your graphs with necessary noise.

# Other background

## Depth contours

In oceanography one is often interested in indicating depth. Global relief models from the ETOPO1 dataset hosted on a NOAA server can be accessed using the `getNOAA.bathy`-function in the marmap-package. To access them one specifies the boundary of the data of interest and then, since we are using ggplot for mapping are turned into a data frame using the `fortify`-function:

```{r}
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)
depth <- 
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                lat1 = ylim[1], lat2 = ylim[2],
                resolution = 1) %>% 
  fortify()  # turn the object into a data.frame
glimpse(depth)
```

So this data is just a set of x (longitude), y (latitudes) and z (depth). The dataset is a raster-grid which we can visualize by using the `geom_raster`-function:

```{r}
depth %>% 
  filter(z <= 0) %>% 
  ggplot() +
  theme_bw() +
  geom_raster(aes(x, y, fill = z)) +
  coord_quickmap(expand = FALSE)
```

We generate the base map with contours as follows:
```{r}
m <- ggplot() +
  theme_bw() +
  geom_contour(data = depth, aes(x, y, z = z),
               breaks=c(-25, -50, -100, -200, -400),
               colour="black", size=0.1) +
  geom_polygon(data = iceland, aes(long, lat, group = group), fill = "grey") +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE) +
  labs(x = NULL, y = NULL)
```

Lets just look at what we have created:
```{r}
m
```

Here we have specified depth contour values of 25, 50, 100, 200 and 400 meters. Now we are ready to add the minke data or any other data of interest:

```{r}
m + geom_point(data = minke, aes(lon, lat), colour = "red")
```


<div class="panel panel-warning">
<div class="panel-heading">Homework</div>
<div class="panel-body">

1. Create a depth raster map of the region of your interest, e.g. for the longitudinal and latitudinal range that covers your data. If you do not have spatial data, use the NS-IBTS station data from the [Working with dates](pre_dates.html) presentation.
2. Create a contour map of the region of your interest, specifying your own preference for the depth range to show.
3. Overlay your own data on top of the contour map.
4. Try a plot where the raster image of depth is the background, overlay the contours and then your data. Although the image may look sexy, think about the **main** message your are trying to convey to the recipient of such a plot.

</div>
</div>


# A case example

## SMB tow tracks

```{r}
con <- connect_mar()
lesa_stodvar(con) %>% 
  filter(synaflokkur == 30,
         ar == 2017) %>% 
  collect(n = Inf) %>% 
  ggplot() +
  geom_polygon(data = island, aes(lon, lat)) +
  geom_segment(aes(x = kastad_v_lengd,
                   y = kastad_n_breidd,
                   xend = hift_v_lengd,
                   yend = hift_n_breidd)) +
  coord_quickmap()
```

## Catch per tow

Lets calculate the catch per tow from the spring survey:

```{r}
d.sql <-
  lesa_stodvar(con) %>% 
  inner_join(lesa_lengdir(con)) %>% 
  filter(tegund == 3,                    # lets look at saithe
         synaflokkur == 30) %>% 
  skala_med_toldum() %>% 
  mutate(b = fjoldi * 0.00001 * lengd^3) %>% 
  rename(lon = kastad_v_lengd, lat = kastad_n_breidd) %>% 
  group_by(synis_id, ar, lon, lat) %>% 
  summarise(n = sum(fjoldi),
            b = sum(b))
d.df <-
  d.sql %>% 
  collect(n = Inf)
d.df %>% 
  ungroup() %>% 
  filter(ar %in% seq(1990, 2015, by = 5)) %>% 
  ggplot(aes(lon, lat)) +
  geom_polygon(data = island) +
  geom_point(aes(size = b),
             colour = "red", alpha = 0.4) +
  scale_size_area(max_size = 20) +
  facet_wrap(~ ar) +
  labs(x = NULL, y = NULL) +
  coord_quickmap()
```

## Gridded data

We are often interested in summarizing spatial data onto a grid for presentations. A typical example are logbook data which often can hold millions of records. Lets read in some bottom trawl logbook data from Iceland:

```{r}
d.sql <-
  afli_stofn(con) %>% 
  left_join(afli_afli(con)) %>%     # catch by species is stored here
  left_join(afli_toga(con)) %>%     # togtimi is stored here
  filter(ar == 2017,
         veidarf == 6) %>%          # bottom trawl
  # "project" coordinates to a 1 x 0.5 degree grid
  #    byproduct: get a new variable name "sq"
  encode_zchords(col.names = c("breidd", "lengd"), dx = 1, dy = 0.5) %>%
  group_by(sq) %>% 
  summarise(catch = sum(afli)/1e3,
            effort = sum(togtimi)/60)
d.df <-
  d.sql %>% 
  collect(n = Inf) %>% 
  separate(col = sq, into = c("lon", "lat"), sep = ":", convert = TRUE)
d.df %>% 
  ggplot() +
  theme_bw() +
  geom_raster(aes(lon, lat, fill = effort/1e3)) +  
  geom_polygon(data = island, aes(lon, lat), fill = "grey") +
  coord_quickmap() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  scale_fill_continuous(type = "viridis", option = "B", direction = -1) +
  labs(fill = "Effort",
       title = "Bottom trawl effort in 1000 hours")
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Play with other species
2. Play with other backgrounds
3. Add layers, e.g. depth contours
4. Try finer grids

</div>
</div>

