---
title: "Spatial data and ggplot"
output: 
  html_document:
    fig_height: 4
    fig_width: 8
    highlight: haddock
    theme: united
    toc: yes
    toc_float: yes
---

```{r, echo = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
Sys.setenv(LANG = "en")
```

# Preamble

Note: Some restructuring of the text, minor edits in the code and some additions were added after the lecture on 2017-01-12.


Libraries needed for this session are:

```{r, message = FALSE, warning = FALSE, results='hide'}
# devtools::install_github("tidyverse/ggplot2")  # has geom_sf
library(tidyverse)
library(lubridate)
library(mar)

# use install.packages("packagename") if packages below are not
#   already on your computer
library(geo)               # the old faithful
library(maps)              # basic low resolution maps
library(mapdata)           # some high resolution maps
library(marmap)            # to get bathymetric data from NOAA or GEBCO
library(ggmap)             # tile backgrounds
library(sp)                # spatial stuff
library(rgdal)             # to read in ESRI files
```

Lets continue using the minke data, if only because it has two columns that refer to geographical coordinates. Lets plot the "kill" location of the minkes: 
```{r}
minke <- read_csv("ftp://ftp.hafro.is/pub/tcrenv/minke.csv")
minke %>% 
  ggplot(aes(lon, lat)) +
  geom_point(colour = "red")
```

In the above plot, we basically have mapped the longitude on the x-axis and the latitude on the y-axis. There are two things missing:

* A background, giving the reader a better indication of the geographical region of the sample location.
* The projection (aspect ratio between the x- and the y-axis) looks wrong

In the geo package there is an object call "island", lets take a peek:
```{r}
glimpse(island)
```
So we have a dataframe of some `r nrow(island)` observations with two columns "lon" and "lat". Lets add this as a background to the minke data and use the "Mercator" projection (the default):

```{r}
minke %>% 
  ggplot(aes(lon, lat)) +
  geom_path(data = geo::island) +   # so, this object is in the geo-package
  geom_point(colour = "red") +
  coord_map(projection = "mercator")
```

We use `geom_path`-function because it respects the order of the data in the dataframe (try to use `geom_line` instead in the above code). The bottom line is that a geographical object is in its simplest term just a dataframe containing longitudinal (x) and latitudinal (y) positions that is **arrange**-ed in a specific way. And the call to `coord_map` just sets the proper aspect ratio between the x- and the y-axis.

And just if it has not been made clear - ggplot works first and foremost, as the rest of the tidyverse, with dataframes. With some few exceptions.

# On backgrounds

## Iceland - example of a shapefile

A popular format for **storing** geometric locations and associated attributes (temperature, chlorophyll, depth, catch, diversity, ...) is [ESRI shapefiles](https://en.wikipedia.org/wiki/Shapefile). The shapefile format can spatially describe vector features: points, lines, and polygons, representing, for example, shorelines, rivers, lakes, eez-boundaries, etc.

The ESRI shapefile format can be read into myriads of software, including R. Here we take an example of a shapefile of the Icelandic shoreline obtained from [LMI](http://www.lmi.is). A thinned version has been conveniently placed at ftp://ftp.hafro.is/pub/reiknid/einar/shapes. To read it into R one would need to download the bundle onto ones computer first:

```{r, eval = FALSE}
dir.create("data")    # creates folder "data" in your working directory if it
                      #  does not already exist.
download.file("ftp://ftp.hafro.is/pub/reiknid/einar/shapes/iceland.zip",
              # store it in the data directory under the filename iceland.zip
              destfile = "data/iceland.zip")
unzip("data/iceland.zip", exdir = "data")
```

Lets see what we have got:
```{r}
dir("data", pattern = "iceland")
```

So we have, besides the .zip file four files ending with .shp, .shx, .dbf and .prj. These are standard extensions of [ESRI shapefiles](https://en.wikipedia.org/wiki/Shapefile) containing needed information on the geometry, indexing, coordinate system and projections.

To read shapefiles into the R we can use the `readOGR` function that resides in the rgdal-package:
```{r}
iceland.sp <- 
  rgdal::readOGR("data/iceland.shp", verbose = FALSE)
class(iceland.sp)
# str(iceland.sp)
```

The class of the object iceland is "SpatialPolygonsDataFrame" and the structure is kind of obtuse. You can plot this as a "base plot" by:

```{r, eval = FALSE}
sp::plot(iceland.sp)
```

ggplot can operate on this class object directly (see below). What really happens behind the scene in ggplot is that the object is **first** turned into a dataframe via the `fortify`-function. Let's look at what that returns:

```{r}
iceland.df <- iceland.sp %>% fortify()
glimpse(iceland.df)
```

When using the `fortify`-function on any spatial object (as defined in the sp-package) you will always get the same variable names: The coordinates names are always "long" and "lat" with each "island" separated by values in the variable "group". Hence to properly render any such objects in ggplot one needs to specify "long" and "lat"  as the x and the y and the group argument in the `aes`-function as the variable name "group":

```{r}
ggplot() +
  geom_polygon(data = iceland.df, aes(long, lat, group = group)) +
  geom_point(data = minke, aes(lon, lat), col = "red") +
  coord_map(xlim = c(-22.8, -21.7), ylim = c(63.95, 64.4))
```

Here one additional element is introduced:

* The limits are specified inside the call in function `coord_quickmap`. This allows us to zoom into a specific region (try `coord_map` without any argument).

If we had not constructed the code as above we would have gotten a bit of a mess. Try:
```{r eval = FALSE}
ggplot() +
  geom_polygon(data = iceland.df, aes(long, lat)) +
  coord_map(xlim = c(-22.8, -21.7), ylim = c(63.95, 64.4))
ggplot() +
  geom_polygon(data = iceland.df, aes(long, lat, group = group)) +
  coord_map() +
  xlim(c(-22.8, -21.7)) + ylim(c(63.95, 64.4))
```

And as said before, if one had passed the spatial object directly to ggplot, it would convert it first "behind" the scene into a dataframe via the fortify-function. Hence we could have skipped that step and done:
```{r, eval = FALSE}
ggplot() +
  geom_polygon(data = iceland.sp, aes(long, lat, group = group)) +
  geom_point(data = minke, aes(lon, lat), col = "red") +
  coord_map(xlim = c(-22.8, -21.7), ylim = c(63.95, 64.4))
```

The original shapefile from LMI, which has a considerable higher resolution is available (at least temporarily) in the "biceland.zip" bundle in the [ftp-directory](ftp://ftp.hafro.is/pub/reiknid/einar/shapes)

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

As an exercise in downloading, extracting, importing and plotting a shapefile in ggplot try to plot the minke data over the bormicon-area shapefile, the location of the zipped bundle being ftp://ftp.hafro.is/pub/reiknid/einar/shapes/bormicon.zip

```{r, echo = FALSE, eval = FALSE}
download.file("ftp://ftp.hafro.is/pub/reiknid/einar/shapes/bormicon.zip",
              destfile = "data/bormicon.zip")
unzip("data/bormicon.zip", exdir = "data")
rgdal::readOGR("data/bormicon.shp", verbose = FALSE) %>% 
  ggplot() +
  geom_path(aes(long, lat, group = group)) +
  geom_point(data = minke, aes(lon, lat), col = "red") +
  coord_quickmap()
```

</div>
</div>

## Global shorelines

Some of you may be working on a larger scale than the some random islands. The country/shorelines accessible in the map-package may be enough to serve your needs. The are turned into a dataframe, suitable for plotting in ggplot with the `map_data`-function

```{r}
m <- map_data("world")
glimpse(m)
ggplot(m, aes(long, lat, group = group)) +
  geom_polygon() +
  coord_quickmap()
```

Although all two-dimensional maps of the globe are wrong, this one definitively is one of the worst :-) But that is not the issue here - just the concept.

A little higher resolution of the global country profiles is available in the mapdata-package. E.g. to get a map of the North Sea one could do:

```{r}
xlim <- c(-5, 12)
ylim <- c(50, 60)
m <- map_data("worldHires", xlim = xlim, ylim = ylim)
glimpse(m)
ggplot(m, aes(long, lat, group = group)) +
  geom_polygon() +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE)
```

#### Exercise

<div class="panel panel-warning">
<div class="panel-heading">Homework</div>
<div class="panel-body">

1. Play around by selecting and plotting different regions or areas
2. Read the `coord_map` help file for projections other than the default "Mercator". Try to create a map with different projections, e.g. one that looks something along (hint: ortho):

```{r, echo = FALSE}
map_data("world") %>% 
  ggplot(aes(x = long, y = lat, group = group)) +
  theme_bw() +
  geom_polygon(fill = "grey") +
  scale_y_continuous(NULL) +
  scale_x_continuous(NULL) +
  coord_map("ortho", xlim = c(-45,30), ylim = c(35,90))
```

</div>
</div>

## Background images

The ggmap-packages allows one to obtain background images from the web via the `get_map`-function:

```{r eval = FALSE}
m2 <- get_map(location = c(-19,65), zoom= 6)
```

```{r echo = FALSE, eval = FALSE}
write_rds(m2, path = "data/google_raster_image.rda")
```

```{r, echo = FALSE}
m2 <- read_rds("data/google_raster_image.rda")
```

```{r}
m2 <- ggmap(m2) +
  labs(x = NULL, y = NULL)
m2
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Read the help file for `get_map` and get at minimum a satellite maptype.
2. Overlay the minke data
3. Try to limit the plot range to the range of the minke data


```{r, echo = FALSE, eval = FALSE}
m2 <- get_map(location = c(-19,65), zoom= 6, maptype = "satellite")
m2 %>% 
  ggmap() +
  geom_point(data = minke, aes(lon, lat), colour = "red") +
  labs(x = NULL, y = NULL) +
  xlim(range(minke$lon)) +    # limit the map range  
  ylim(range(minke$lat))      # to the data range
```

</div>
</div>

Although using background "images" are impressive, the novelty wears off quickly. Think carefully about not cluttering your graphs with superfluous noise.

# Other background

## Raster maps

In oceanography one is often interested in indicating depth. Global relief models from the ETOPO1 dataset hosted on a NOAA server can be accessed using the `getNOAA.bathy`-function in the marmap-package. To access them one specifies the boundary of the data of interest and then, since we are using ggplot for mapping are turned into a data frame using the `fortify`-function:

```{r}
xlim <- c(-28, -10)
ylim <- c(62.5, 67.5)
depth <- 
  getNOAA.bathy(lon1 = xlim[1], lon2 = xlim[2],
                lat1 = ylim[1], lat2 = ylim[2],
                resolution = 1) %>% 
  fortify()  # turn the object into a data.frame
glimpse(depth)
```

So this data is just a set of regular spaced x (longitude) and y (latitudes) coordinates with z (depth) as an attribute within each grid cell. This kind of data one can visualize using the `geom_raster`-function:

```{r}
depth %>% 
  filter(z <= 0) %>% 
  ggplot() +
  theme_bw() +
  geom_raster(aes(x, y, fill = z)) +
  coord_quickmap(expand = FALSE)
```

## Contours

If we have a regular raster object one can generate a simple contour map from them within ggplot as follows:
```{r}
m <- ggplot() +
  theme_bw() +
  geom_contour(data = depth, aes(x, y, z = z),
               breaks=c(-25, -50, -100, -200, -400),
               colour="black", size=0.1) +
  geom_polygon(data = iceland.df, aes(long, lat, group = group), fill = "grey") +
  coord_quickmap(xlim = xlim, ylim = ylim, expand = FALSE) +
  labs(x = NULL, y = NULL)
```

Lets just look at what we have created:
```{r}
m
```

Here we have specified depth contour values of 25, 50, 100, 200 and 400 meters. Now we are ready to add the minke data or any other data of interest:

```{r}
m + geom_point(data = minke, aes(lon, lat), colour = "red")
```


<div class="panel panel-warning">
<div class="panel-heading">Homework</div>
<div class="panel-body">

1. Create a depth raster map of the region of your interest, e.g. for the longitudinal and latitudinal range that covers your data.
2. Create a contour map of the region of your interest, specifying your own preference for the depth contour lines to show.
3. Overlay your own data on top of the contour map.
4. Try a plot where the raster image of depth is the background, overlay the contours and then your data. Although the image may look sexy, think about the **main** message your are trying to convey to the recipient of such a plot.

</div>
</div>


# Simple case examples

Following examples show how to generate the most common types of gis visual representations used at MFRI in ggplot.

## Tow tracks

```{r}
con <- connect_mar()
lesa_stodvar(con) %>% 
  # Spring bottom trawl survey 2017
  filter(synaflokkur == 30,
         ar == 2017) %>% 
  collect(n = Inf) %>% 
  ggplot() +
  geom_polygon(data = island, aes(lon, lat)) +
  geom_segment(aes(x = kastad_v_lengd,
                   y = kastad_n_breidd,
                   xend = hift_v_lengd,
                   yend = hift_n_breidd)) +
  coord_quickmap()
```

## Cruise tracks

Cruise tracks are something we are often interested to present. Here we take an example of the cruise tracks of Árni and Bjarni in 2017.

These data, among other ship tracks are stored in schema 'stk', that has restricted access. For those that have access one can extract the tracks as follows (for those that do not, skip this step):

```{r, eval = FALSE}
track <-
  tbl_mar(con, "stk.stk_vms_v") %>%
  mutate(ar = to_char(posdate, "YYYY")) %>%
  filter(ar == "2017",
         skip_nr %in% c("1131", "2350")) %>%
  select(-ar) %>% 
  collect(n = Inf)
write_csv(track, path = "/net/www/export/home/ftp/pub/tcrenv/track.csv")
```

Lets read in the tracks from the ftp-site
```{r}
track <- read_csv("/net/www/export/home/ftp/pub/tcrenv/track.csv")
glimpse(track)
```

So we have some `r nrow(track)` recordings. The variable of interest are:

* skip_nr: The ship id in the national registry
* posdate: The date and time of the position recording
* poslon: The x-position
* poslat: The y-position

The poslon and poslat are in a xxx-coordinate system which we first need to convert to our conventional one using a formula from god:
```{r}
track <-
  track %>% 
  mutate(poslon = poslon * 45 / atan(1),
         poslat = poslat * 45 / atan(1)) %>%
  arrange(skip_nr, posdate) %>%
  # The capelin cruise went to the Carribean
  filter(poslon > -50)
```

Take note that the data was arranged by skip_nr and then posdate - this to ensure a proper order of the data once passed to ggplot below:

One can get a view of the data using the following code, but it lacks a background reference:
```{r eval = FALSE}
track %>% 
  ggplot() +
  geom_path(aes(poslon, poslat, colour = as.character(skip_nr))) +
  labs(colour = "Ship") +
  coord_quickmap()
```

A background reference can generate e.g. by:
```{r}
m <- map_data("world") %>% 
  ggplot() +
  theme_bw() +
  geom_polygon(aes(x = long, y = lat, group = group), fill = "grey") +
  scale_y_continuous(NULL) +
  scale_x_continuous(NULL) +
  coord_map("ortho", xlim = c(-45, 25), ylim = c(60, 80))
```

And then overlaying the track data would simply be:
```{r}
m +
  geom_path(data = track, aes(poslon, poslat, colour = as.character(skip_nr))) +
  labs(colour = "Ship") +
  theme(legend.position = c(0.2, 0.7)) +
  scale_color_brewer(palette = "Set1")
```

We note that there are some "erratic" jumps in the tracks, e.g. around "Svalbard" that may need to be fixed, but lets leave that for now.

## Something per tow

Lets calculate the catch per tow from the spring survey:

```{r}
d.sql <-
  lesa_stodvar(con) %>% 
  inner_join(lesa_lengdir(con)) %>% 
  filter(tegund == 3,                    # lets look at saithe
         synaflokkur == 30) %>% 
  skala_med_toldum() %>% 
  mutate(b = fjoldi * 0.00001 * lengd^3) %>% 
  rename(lon = kastad_v_lengd, lat = kastad_n_breidd) %>% 
  group_by(synis_id, ar, lon, lat) %>% 
  summarise(n = sum(fjoldi),
            b = sum(b))
d.df <-
  d.sql %>% 
  collect(n = Inf)
d.df %>% 
  ungroup() %>% 
  filter(ar %in% seq(1990, 2015, by = 5)) %>% 
  ggplot(aes(lon, lat)) +
  geom_polygon(data = island) +
  geom_point(aes(size = b),
             colour = "red", alpha = 0.4) +
  scale_size_area(max_size = 20) +
  facet_wrap(~ ar) +
  labs(x = NULL, y = NULL) +
  coord_quickmap()
```

## Gridding data

We are often interested in summarizing spatial data onto a grid for presentations. A typical example are logbook data which often can hold tens of thousand of records. Lets read in some bottom trawl logbook data from Iceland:

```{r}
d.sql <-
  afli_stofn(con) %>% 
  left_join(afli_afli(con)) %>%     # catch by species is stored here
  left_join(afli_toga(con)) %>%     # togtimi is stored here
  filter(ar == 2017,
         veidarf == 6) %>%          # bottom trawl
  # "project" coordinates to a 1 x 0.5 degree grid
  #    byproduct: get a new variable name "sq"
  encode_zchords(col.names = c("breidd", "lengd"), dx = 1, dy = 0.5) %>%
  group_by(sq) %>% 
  summarise(catch = sum(afli)/1e3,
            effort = sum(togtimi)/60)
d.df <-
  d.sql %>% 
  collect(n = Inf) %>% 
  separate(col = sq, into = c("lon", "lat"), sep = ":", convert = TRUE) %>% 
  # a temporary fix for windose because in the above the lon and the lat
  #   are not converted to numeric values.
  # so here, replace "," with "." in the lon and lat, then convert to numeric
  mutate(lon = gsub(',','.',lon) %>% as.numeric(),
         lat = gsub(',','.',lat) %>% as.numeric())
d.df %>% 
  ggplot() +
  theme_bw() +
  geom_raster(aes(lon, lat, fill = effort/1e3)) +  
  geom_polygon(data = island, aes(lon, lat), fill = "grey") +
  coord_quickmap() +
  scale_x_continuous(name = NULL, breaks = NULL) +
  scale_y_continuous(name = NULL, breaks = NULL) +
  scale_fill_viridis_c(option = "B", direction = -1) +
  labs(fill = "Effort",
       title = "Bottom trawl effort in 1000 hours")
```

<div class="panel panel-warning">
<div class="panel-heading">Exercise</div>
<div class="panel-body">

1. Play with other species
2. Play with other backgrounds
3. Add layers, e.g. depth contours
4. Try finer grids

</div>
</div>
