---
title: "Solutions"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r, echo = FALSE}
library(tidyverse)
```

# Getting data into R

<div class="panel panel-warning">
<div class="panel-heading">Exercise 1</div>
<div class="panel-body">

1. How would you import data where the column separator is "|"

```{r, eval = FALSE}
read_delim(file = "filename",
           delim = "|")
```


2. You current working directory is "~/Documents/workdir1/" and you want to read a file named "data.csv" in "~/Documents/datadir/". Can you think of two ways to write the file path?

```{r, eval = FALSE}
read_csv(file = "~/Documents/datadir/data.csv")
read_csv(file = "../datadir/data.csv")
```


3. Copy the following text to a text file and read in this file using `read_delim`:
```
this is some data
col1 col2 col3
1 2 a
3 4 b
5 6 c
```

```{r, eval = FALSE}
read_delim("this is some data
col1 col2 col3
1 2 a
3 4 b
5 6 c",
delim = " ",
skip = 1)
```


</div>
</div>

<div class="panel panel-warning">
<div class="panel-heading">Exercise 2</div>
<div class="panel-body">


1. What is wrong with this file
```
a,b,c 
1,1,bla,bla
2,2,bingo
```
Second line has four variables but column names (1st line) and variables in line 3 are only three.

Hence we get:
```{r}
read_csv("a,b,c 
1,1,bla,bla
2,2,bingo")
```




2. What went wrong with the import?
```{r, echo=FALSE, message=FALSE, warning=FALSE}
tmp <- read_table('data/minke.csv')
head(tmp)
```

```{r, eval = FALSE}
tmp <- read_table('data/minke.csv')
head(tmp)
```

</div>
</div>

